---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(whyna)
```

# whyna: Informative Missingness via vctrs

<!-- badges: start -->
<!-- badges: end -->

**whyna** implements typed/informative missingness in R, letting you track *why* values are missing. Standard R treats all missing values as identical `NA`s, losing critical information about the missingness mechanism. This package introduces a custom vector class that stores both the observed value (or `NA`) and the reason it's missing, following Rubin's (1976) MCAR/MAR/MNAR framework.

## Installation

```r
# Install from GitHub:
devtools::install_github("brownag/whyna")
```

## Why Informative Missingness?

In real data, missing values arise from different mechanisms, each with distinct mathematical and inferential properties. Rubin's (1976) taxonomy formalizes these via the conditional probability of missingness given the complete data.

**MCAR** (Missing Completely At Random)
: Missingness is independent of both observed and unobserved data.
  Safe to delete; standard listwise deletion yields unbiased estimates.

**MAR** (Missing At Random)
: Missingness depends on observed data but not the unobserved value itself.
  The mechanism is ignorable: valid inference requires only maximum likelihood or multiple imputation (MI) methods, without explicitly modelling the missingness process.

**MNAR** (Missing Not At Random)
: Missingness depends on the unobserved value itself.
  The mechanism is non-ignorable: standard analysis methods introduce severe bias. Sensitivity analysis or specialized methods (selection models, pattern-mixture models) are required.

### Ignorability and Statistical Inference

A missingness mechanism is ignorable if the analyst can obtain valid parameter estimates without explicitly modelling the missingness process. This holds if and only if:
1. The data are MAR, and
2. The missingness parameters are distinct from the data-generating parameters (no cross-constraints).

Under ignorability, standard likelihood-based inference proceeds normally. MNAR violates ignorability, forcing analysts to model the missingness mechanism directly.

### A Critical Caveat: Three-Valued Logic

R uses three-valued logic: TRUE, FALSE, and NA (unknown). When NA appears in a comparison, the result is NA because the truth value cannot be determined.

```r
NA > 65          # returns NA, not TRUE or FALSE
is.na(NA > 65)   # TRUE
```

This is correct behavior. The alternative—treating NA as an extreme value (smallest or largest number)—produces silent errors. If NA were treated as the maximum value, then `age > 65` would incorrectly include all missing ages in the result set. If NA were the minimum value, it would exclude them. The result silently reports something untrue.

R and whyna preserve three-valued logic. When you filter or compare, NA propagates naturally, and you must explicitly handle it with `is.na()` or `na.rm = TRUE`. This forces you to confront missing data rather than accidentally corrupting your analysis.

## Quick Start

```{r quick-start}
library(whyna)

measurements <- why_na(
  c(5.2, NA, 7.1, NA, 3.8),
  type = c(0L, 1L, 0L, 2L, 3L)  # 0=Observed, 1=MCAR, 2=MAR, 3=MNAR
)

measurements

# Extract missingness predicates
is_mcar(measurements)

is_mar(measurements)

is_mnar(measurements)

# Get a summary factor
missing_reason(measurements)
```

## Generalized Typed NA: Beyond Numeric Data

**whyna** now supports any vector type, not just doubles. Here's an example with character data where sentinel codes (like "LOD" for "limit of detection") and normal values coexist:

```{r typed-na-character}
# Character example: lab measurement codes
lab_results <- c("42.5", "LOD", "38.2", "NULL-broken", "45.1")
measurements <- sentinel_recode(
  lab_results,
  map = list("LOD" = "MNAR", "NULL-broken" = "MCAR"),
  ptype = character()
)

measurements
missing_reason(measurements)
```

Integer example with type preservation:

```{r typed-na-integer}
counts <- why_na(c(10L, NA, 5L, NA, 8L), c(0L, 1L, 0L, 2L, 0L))
counts  # Type is preserved: integer, not double
```

## Realistic Case: Water Quality Monitoring

Suppose you're analyzing dissolved oxygen (DO) measurements from 50 water samples:

```{r water-data-setup}
library(whyna)
library(tibble)
library(dplyr, warn.conflicts = FALSE)

# Simulated water quality dataset
set.seed(123)
n <- 50

water_data <- tibble(
  site_id = 1:n,
  do_value = c(rnorm(35, mean = 8, sd = 1), rep(NA_real_, 15)),
  do_type = c(rep(0L, 35), rep(1L, 5), rep(3L, 10)),
  site_type = sample(c("upstream", "downstream"), n, replace = TRUE),
  temperature = rnorm(n, mean = 15, sd = 3),
  ph = rnorm(n, mean = 7.5, sd = 0.5)
)

# Now create the typed missing-data vector
water_data <- water_data |>
  mutate(
    do = why_na(do_value, type = do_type)
  ) |>
  select(site_id, do, site_type, temperature, ph)

water_data
```

### Analysis Workflow

**1. Describe missingness patterns**

```{r analysis-1}
water_data |>
  summarise(
    n_obs = sum(!is.na(do)),
    n_mcar = sum(is_mcar(do)),
    n_mar = sum(is_mar(do)),
    n_mnar = sum(is_mnar(do))
  )
```

**2. Filter by missingness mechanism for sensitivity analysis**

```{r analysis-2}
# Rows with observed data only
observed <- water_data |> filter(!is.na(do))
nrow(observed)

# Rows with observed or MCAR (safe to use)
trustworthy <- water_data |> filter(!is.na(do) | is_mcar(do))
nrow(trustworthy)

# Check for MNAR (biasing mechanism)
any(is_mnar(water_data$do))
```

**3. Stratified analysis by site type**

```{r analysis-3}
water_data |>
  group_by(site_type) |>
  summarise(
    n = n(),
    mean_do = mean(as.double(do), na.rm = TRUE),
    pct_mnar = 100 * mean(is_mnar(do)),
    .groups = "drop"
  )
```

**4. Extract the complete-case subset (removes all missing, regardless of type)**

```{r analysis-4}
complete <- water_data |> filter(!is.na(do))
nrow(complete)
head(complete)
```

## API Reference

### Constructors

**`why_na(x, type = NULL)`**
User-facing constructor. Auto-detects missingness if `type` omitted.

```{r api-constructor-1}
why_na(c(1.5, NA, 2.3))
```

**`informative_na(x, type = NULL)`**
Alias for `why_na()` (verbose form, identical behavior).

**`new_informative_na(value, type)`**
Low-level constructor; strict validation, for developers.

### Extraction & Predicates

**`is_mcar(x)`, `is_mar(x)`, `is_mnar(x)`**
Logical masks indicating which elements have a given missingness type.

```{r api-predicates}
x <- why_na(c(1, NA, 3, NA), c(0L, 1L, 0L, 2L))
is_mcar(x)
is_mar(x)
```

**`missing_reason(x)`**
Returns a factor with levels: `Observed`, `MCAR`, `MAR`, `MNAR`.

```{r api-reason}
missing_reason(x)
```

**`is_observed(x)`, `is_ignorable(x)`**
Additional predicates for analysis workflows.

```{r api-additional-predicates}
is_observed(x)
is_ignorable(x)
```

**`missing_summary(x)`**
Quick overview of missingness patterns: returns a data.frame with mechanism counts and proportions.

```{r api-missing-summary}
missing_summary(x)
```

### Data Ingestion

**`sentinel_recode(x, map)`**
Maps domain-specific sentinel codes (e.g., `"-9999"`, `"LOD"`, `"NULL-broken"`) to a whyna vector.
Numeric values pass through as observed.

```{r api-sentinel-recode}
raw <- c("12.5", "-9999", "LOD", "15.1")
x <- sentinel_recode(raw, list("-9999" = "MAR", "LOD" = "MNAR"))
x
```

### Base R Integration

**`is.na(x)`**
Returns `TRUE` for all missing entries (regardless of type).

**`format(x)`, `print(x)`**
Displays observed values numerically and missing as `<NA:MCAR>`, etc.

**`c()`, `rbind()`**
Combines with plain numeric vectors, upgrading to `informative_na`.

```{r api-combine}
x <- why_na(c(1.5, NA), c(0L, 1L))
y <- c(3.1, 4.2)
c(x, y)
```

**Cast to `double`**
Extracts the value field (lossy; discards type information).

```{r api-cast}
as.double(x)
```

## Design Notes

- **Immutability of types**: When `type != 0L`, the `value` field is always `NA` of the appropriate type (e.g., `NA_integer_` for integer vectors, `NA_character_` for character, `<NA>` for factors).
- **Type preservation**: Input type is preserved in `why_na()`. Use the optional `ptype` argument for explicit casting.
- **vctrs foundation**: Built on `vctrs::new_rcrd()` for integration with tidyverse (dplyr, tibble).
- **No arithmetic**: Operations like `x + 1` return standard `NA` (not class-preserving). Types are preserved during subsetting but revert to standard NA under arithmetic operations.

## References

**Foundational Theory**

- Rubin, D. B. (1976). Inference and missing data. *Biometrika*, 63(3), 581–592.
- Little, R. J. A., & Rubin, D. B. (2002). *Statistical analysis with missing data* (2nd ed.). Wiley.

**Practical Methods**

- Carpenter, J. R., & Kenward, M. G. (2013). *Multiple imputation and its application*. Wiley.
- van Buuren, S. (2018). *Flexible imputation of missing data* (2nd ed.). CRC Press.
- Heckman, J. J. (1979). Sample selection bias as a specification error. *Econometrica*, 47(1), 153–161.

## License

MIT License. See LICENSE file for details.

---

**Questions?** File an issue or open a discussion. Contributions welcome!
